---
title: "Shiny Workshop"
output: 
  learnr::tutorial:
    progressive: false
    allow_skip: true
    df_print: default
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

if(!("ggplot2" %in% rownames(installed.packages()))) {
  install.packages("ggplot2")
}

library(learnr)
library(tidyverse)
library(knitr)
library(shiny)
```

## Introduction

 - Mike Kleinsasser, MS in Statistics

 - R programmer for the Department of Biostatistics
 
 - Consult faculty and students on R package development, optimization
 
 - Give workshops to students on useful R tools
   - First workshops this semester: Tidyverse Series

 - Data Manipulation Workshop: Tuesday, September 15 at 5 p.m.
   - Zoom link: https://umich.zoom.us/j/98858754267
   - RSVP: https://www.eventbrite.com/e/data-manipulation-workshop-tickets-118357150423
 - Data Visualization Workshop: Thursday, September 24 at 5 p.m.
    - Zoom link: https://umich.zoom.us/j/98858754267
    - RSVP: https://www.eventbrite.com/e/data-visualization-workshop-tickets-118360271759
 - Shiny Workshop: Tuesday, October 6 at 5 p.m.
    - Zoom link: https://umich.zoom.us/j/98858754267
    - RSVP: https://www.eventbrite.com/e/shiny-workshop-tickets-118360693019
 - Package Development 1: Thursday, October 15 at 5 p.m.
    - Zoom link: https://umich.zoom.us/j/98858754267
    - RSVP: https://www.eventbrite.com/e/package-development-1-tickets-118360891613
 - Package Development 2: Tuesday, October 27 at 5 p.m.
    - Zoom link: https://umich.zoom.us/j/98858754267
    - RSVP: https://www.eventbrite.com/e/package-development-2-tickets-118361158411
    
Zoom join information:
Mike Kleinsasser is inviting you to a scheduled Zoom meeting.

Topic: R Workshops (Recurring)

Join Zoom Meeting
https://umich.zoom.us/j/98858754267

Meeting ID: 988 5875 4267
One tap mobile
+13017158592,,98858754267# US (Germantown)
+13126266799,,98858754267# US (Chicago)

Dial by your location
        +1 301 715 8592 US (Germantown)
        +1 312 626 6799 US (Chicago)
        +1 646 876 9923 US (New York)
        +1 253 215 8782 US (Tacoma)
        +1 346 248 7799 US (Houston)
        +1 669 900 6833 US (San Jose)
        +1 647 374 4685 Canada
        +1 647 558 0588 Canada
        +1 778 907 2071 Canada
        +1 204 272 7920 Canada
        +1 438 809 7799 Canada
        +1 587 328 1099 Canada
Meeting ID: 988 5875 4267
Find your local number: https://umich.zoom.us/u/acTYHlRYm

## Introduction to Shiny

 - Install R package `Shiny` from CRAN: `install.packages("shiny")`
  
 - `Shiny` makes it easy to build web applications from R
 
 - Package has 11 build-in examples showing off diverse features
 
```{r e1, exercise=FALSE, eval=FALSE, echo=TRUE}
library(shiny)
runExample("01_hello")
```

 - you're expected to have an R session running with Shiny installed for this session
 
 - also, install `quantmod` from CRAN for our extended example
 
## What are the components of an app?

 - Apps are made of two functions:
 
 - 1. the server
 
 - server is the engine of the app; it does the heavy lifting, e.g., computing the regression coefficients for a user specified model
 
 - the server is implemented as a function with instructions on how to build and 
 rebuild the R objects displayed in the UI
 
 - 2. the UI (user interface)

 - the UI is what the user interacts with through keyboard and mouse input, .e.g., upload a data set to fit by opening a file explorer
 
 - also a function, but implemented as a set of nested R functions that assemble 
 an HTML user interface for your app
 
 - 3. and, a call to `shinyApp` function which simply takes the server and UI
 
 - step 3. creates a Shiny app object from an explicit server/UI pair
 
 - `shinyApp` combines the ui and server into a functional app
 
## Running an app

 - R session will be busy while the app is running
 
 - cannot run other R commands, why?
 
 - Since web apps are interactive, R needs to monitor the UI for changes and
  execute the app's reactions (more on this later)
 
 - to kill the app, hit escape or click the stop sign icon
 
 - A first app: (copy the code into your R session)
 
```{r, exercise=FALSE, eval=FALSE, echo=TRUE}
library(shiny) 
ui <- fluidPage() 
server <- function(input, output){} 
shinyApp(ui = ui, server = server)
```

 - The app is blank!
 
 - print ui the function, notice that it's just a wrapper for html
 
```{r, exercise=FALSE, eval=FALSE, echo=TRUE}
ui
```

 - Note: the easiest way to share app is through shinyapps.io, a cloud based service from RStudio. The department has a standard plan, which is available for use by students and faculty

## Reactive output

 - the idea of a web app is to have a "live" quality
 
 - reactive output is output that automatically responds when user toggles a widget
 
 - Step 1. Add an R object to the UI
 
```{r, exercise=FALSE, eval=FALSE, echo=TRUE}
# Example, plot the R object called hist
ui = fluidPage(
  plotOutput("hist")
)
```
 
 - Step 2. Instruct Shiny how to build the object from the server function. 
 
```{r, exercise=FALSE, eval=FALSE, echo=TRUE}
server = function(input, output) {
  # update the R object called hist based on widget value
  output$hist = renderPlot({
    hist(input$data)
  })
}
```
 
 - the object is reactive if code that builds it calls a widget value

 - this back and forth process creates the live quality
 
 - Add inputs to the UI with *Input() functions
 
 - Add outputs to the UI with the *Output() functions
 
 - Then, in the server, tell it how to render the output with:
 
 - render*() functions

## build the UI

 - First, lay out the user interface
 
 - add text, images, buttons, text boxes, etc.
 
```{r, exercise=FALSE, eval=FALSE, echo=TRUE}
library(shiny)

# Define UI ----
ui <- fluidPage(
  
)

# Define server logic ----
server <- function(input, output) {
  
}

# Run the app ----
shinyApp(ui = ui, server = server)
```

 - function `fluidPage` used to display a window that adjusts fluidly, so this
 forms the basis of the UI
 
 - to the fluid page below, add:
 
 - 1. titlePanel with text
 
 - 2. sidebarLayout, with: sidebarPanel and mainPanel (each with text)
 
```{r, exercise=FALSE, eval=FALSE, echo=TRUE}
ui = fluidPage()
```

## add control widgets

 - a widget is a web element that users interact with
 
 - Shiny widgets collect values from user and send them to the server
 
 - Shiny comes with a family of widgets, but add on packages will give you
 some more flexibility
 
 - basic widget examples:
 
 - buttons
 
```{r, exercise=FALSE, eval=FALSE, echo=TRUE}
actionButton()
```

 - checkboxes
 
```{r, exercise=FALSE, eval=FALSE, echo=TRUE}
checkboxInput()
checkboxGroupInput()
```
 
 - numeric input
 
```{r, exercise=FALSE, eval=FALSE, echo=TRUE}
numericInput()
```
 
 - file input
 
```{r, exercise=FALSE, eval=FALSE, echo=TRUE}
fileInput()
```
 
 - select box
 
```{r, exercise=FALSE, eval=FALSE, echo=TRUE}
selectInput()
```
 
 - help text
 
```{r, exercise=FALSE, eval=FALSE, echo=TRUE}
helpText()
```
 
## Add R objects to UI

 - family of functions to turn R objects into output for user interface
 
 - examples:
 
 - to render a data table,
 
```{r, exercise=FALSE, eval=FALSE, echo=TRUE}
dataTableOutput()
```

 - image
 
```{r, exercise=FALSE, eval=FALSE, echo=TRUE}
imageOutput()
```

 - plot
 
```{r, exercise=FALSE, eval=FALSE, echo=TRUE}
plotOutput()
```

 - table
 
```{r, exercise=FALSE, eval=FALSE, echo=TRUE}
tableOutput()
```

## Examples

```{r, exercise=FALSE, eval=FALSE, echo=TRUE}
# install packages ----
install.packages(c('shiny', 'quantmod'))

# Load packages ----
library(shiny)
library(quantmod)
```

 - Start with the basic app template:

```{r, exercise=FALSE, eval=FALSE, echo=TRUE}
# basic app ---- copy from cheat sheet
library(shiny) 
ui <- fluidPage() 
server <- function(input, output){} 
shinyApp(ui = ui, server = server)
```

 - Example 1. First, let's build a simple app that takes a numeric input n (`numericInput`), and draws n random normal samples. Then, display a histogram of the draws by rendering the plot (`renderPlot`)
 
```{r, exercise=FALSE, eval=FALSE, echo=TRUE}
ui = fluidPage(
  
)
server = function(input, output) {
  
}
shinyApp(ui, server)
```

 - Example 2. Real time stock ticker app
 
 - Starting with the UI for a stock visualizer app, 
 
 - First, run it and assess the server components needed,
 
 - Second, analyze the server:
 
```{r, exercise=FALSE, eval=FALSE, echo=TRUE}
library(quantmod)
# Assembled UI
ui = fluidPage(
  titlePanel('Stock Vis'),
  sidebarLayout(
    sidebarPanel(
      textInput('symb', 'Symbol', 'MSFT'),
      dateRangeInput('dates', 'Date Range', start = '2018-01-01', end = as.character(Sys.Date())),
      p(strong('Scale')),
      checkboxInput('log', 'Plot y on the log scale', value = FALSE),
      checkboxInput('adjust', 'Adjust prices for inflation', value = FALSE)
    ),
    mainPanel(
      plotOutput('plot')
    )
  )
)
server = function(input, output) {
}
shinyApp(ui, server)
```

```{r, exercise=FALSE, eval=FALSE, echo=TRUE}
# Server logic ----
server <- function(input, output) {
  
  if (!exists(".inflation")) {
  .inflation <- getSymbols('CPIAUCNS', src = 'FRED', 
     auto.assign = FALSE)
  }  

# adjusts Google finance data with the monthly consumer price index 
# values provided by the Federal Reserve of St. Louis
# historical prices are returned in present values 
  adjust <- function(data) {

      latestcpi <- last(.inflation)[[1]]
      inf.latest <- time(last(.inflation))
      months <- split(data)               
      
      adjust_month <- function(month) {               
        date <- substr(min(time(month[1]), inf.latest), 1, 7)
        coredata(month) * latestcpi / .inflation[date][[1]]
      }
      
      adjs <- lapply(months, adjust_month)
      adj <- do.call("rbind", adjs)
      axts <- xts(adj, order.by = time(data))
      axts[ , 5] <- Vo(data)
      axts
  }    
  
  # grab data
  # store it
  dataInput = reactive({
    req(input$symb != '')
    getSymbols(input$symb, scr = 'yahoo',
               from = input$dates[1], to = input$dates[2],
               auto.assign = FALSE)
  })
  
  # adjusts for inflation
  finalInput = reactive({
    if(!input$adjust) {
      return(dataInput())
    }
    adjust(dataInput())
  })
  
  output$plot = renderPlot({
    chartSeries(finalInput(), name = input$symb, theme = chartTheme('white'), 
                type = 'line', log.scale = FALSE, TA = NULL)
  })
}

# Run the app
shinyApp(ui, server)
```

 - Exercise: Modify the app to allow the user to select log scale for the plot,
 currently the default is TRUE
 
